Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-06-11 paper/analysis.tex
added a few references to the CBJ paper.

We extended our analysis to fainter magnitude limits by replacing Gaia's magnitudes based on trigonometric distance with those based on
the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2
magnitudes deeper and also extends to low halo-like metallicity range. The scatter between photometric $M_r$ based on SDSS data and
$M_r$ based on ``photogeometric'' distances is 0.38 mag. As shown in Figure~\ref{fig:gaiasdss2}, the systematic error
increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available
(both using Gaia data and globular clusters).

>>>>

We extended our analysis to fainter magnitude limits by replacing Gaia's magnitudes based on trigonometric distance with those based on
the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2
magnitudes deeper and also extends to low halo-like metallicity range. The scatter between photometric $M_r$ based on SDSS data and
$M_r$ based on ``photogeometric'' distances is 0.38 mag. As shown in Figure~\ref{fig:gaiasdss2}, the systematic error
increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear\footnote{The so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021} do not utilize [Fe/H] information. Since the [Fe/H] distribution varies with apparent magnitude, one might expect biased distances (20\% or more for halo vs. disk comparison). We find no significant bias compared to our estimates, which probably indicates that their distance priors absorb the impact of unknown [Fe/H]. Hence, the only net effect of unknown [Fe/H] is increased uncertainty of estimated distances (for halo stars, with a metallicity scatter of 0.3 dex, the expected contribution to $M_r$ scatter is about 0.2 mag).}, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available
(both using Gaia data and globular clusters).

### 2024-06-11 paper/methodology.tex
added a few references to the CBJ paper.

The observed stellar colors also depend on the interstellar dust extinction along the line of sight. Hereafter
we will assume (and justify further below) that dust extinction is fully specified by a single model parameter, $A_r$.
$A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, that is, the
posterior probability distribution $q(Q_r)$, where $Q_r=M_r+A_r$ is known, the distance modulus $\Delta$ can be
computed using relationship
\begin{equation}
  \label{eq:distmod}
                 r = M_r + A_r + \Delta.
\end{equation}
In practice the uncertainty of the observed magnitude, $r^{obs}$, is always much smaller than the width of $q(Q_r)$
distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag). Therefore,
\begin{equation}
  \label{eq:distmodpdf}
                         p(\Delta) = q(r^{obs} - \Delta),
\end{equation}
with the mean values related as $\mean{\Delta}$ = $r^{obs} - \mean{Q_r}$ and the uncertainty of $\Delta$
approximately equal to the standard deviation of $q(Q_r)$.

@@

The prior distribution of stellar model parameters, $p(M_r, [Fe/H])$, depends on sky position and apparent
magnitude (here $r$) due to the complex structure of the Milky Way. For example,  \cite{2014ApJ...783..114G}
generated priors using SDSS-based analytic descriptions of the 3-dimensional stellar distribution in the Milky Way
\citep{2008ApJ...673..864J} and metallicity distribution for disk and halo components \citep{2008ApJ...684..287I}. 
Since LSST aims to provide good coverage of the Galactic plane and the Bulge, where SDSS data did not provide strong
constraints, we opted to utilize recent simulations that implement constraints from a variety of modern surveys (some
include the Galactic plane and Bulge regions).

@@

Figure~\ref{fig:priorsAstroLab} shows examples of such $p(M_r, [Fe/H])$ maps for a grid of apparent magnitudes and for
several characteristic sky positions. Several features are noticeable: i) there are two ``clouds'', at low and high metallicity,
that correspond to halo and disk stars; ii) the distribution towards the Galactic center (top row) is much more compact in
the $M_r$ direction (because these subsamples are dominated by the Bulge stars at similar distances); iii) the whole 
distribution shifts to fainter levels as the bin center ($r$) becomes fainter. We have experimented with different region
sizes\footnote{We use HEALPix geometry, see https://healpix.jpl.nasa.gov.} and found that $\sim$10 deg$^2$
(HEALPix $N_{side}$=16 and 32, with 3,072 and 12,288 pixels over the full sky) is sufficiently small to capture the
variation of $p(M_r, [Fe/H])$ maps across the sky.

>>>>

The observed stellar colors also depend on the interstellar dust extinction along the line of sight. Hereafter
we will assume (and justify further below) that dust extinction is fully specified by a single model parameter, $A_r$.
$A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, that is, the
posterior probability distribution $q(Q_r)$, where $Q_r=M_r+A_r$ is known, the distance modulus $\Delta$ can be
computed using relationship
\begin{equation}
  \label{eq:distmod}
                 r = M_r + A_r + \Delta.
\end{equation}
In practice the uncertainty of the observed magnitude, $r^{obs}$, is always much smaller than the width of $q(Q_r)$
distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag). Therefore,
\begin{equation}
  \label{eq:distmodpdf}
                         p(\Delta) = q(r^{obs} - \Delta),
\end{equation}
with the mean values related as $\mean{\Delta}$ = $r^{obs} - \mean{Q_r}$ and the uncertainty of $\Delta$
approximately equal to the standard deviation of $q(Q_r)$. For related discussion, please see \S2.4 in \cite{bailer-jones_estimating_2021}.

@@

The prior distribution of stellar model parameters, $p(M_r, [Fe/H])$, depends on sky position and apparent
magnitude (here $r$) due to the complex structure of the Milky Way. For example,  \cite{2014ApJ...783..114G}
generated priors using SDSS-based analytic descriptions of the 3-dimensional stellar distribution in the Milky Way
\citep{2008ApJ...673..864J} and metallicity distribution for disk and halo components \citep{2008ApJ...684..287I}. 
Since LSST aims to provide good coverage of the Galactic plane and the Bulge, where SDSS data did not provide strong
constraints, we opted to utilize recent simulations that implement constraints from a variety of modern surveys (some
include the Galactic plane and Bulge regions). An analogous approach was taken by \cite{bailer-jones_estimating_2021}.

@@

Figure~\ref{fig:priorsAstroLab} shows examples of such $p(M_r, [Fe/H])$ maps for a grid of apparent magnitudes and for
several characteristic sky positions. Several features are noticeable: i) there are two ``clouds'', at low and high metallicity,
that correspond to halo and disk stars; ii) the distribution towards the Galactic center (top row) is much more compact in
the $M_r$ direction (because these subsamples are dominated by the Bulge stars at similar distances); iii) the whole 
distribution shifts to fainter levels as the bin center ($r$) becomes fainter. We have experimented with different region
sizes\footnote{Following \cite{bailer-jones_estimating_2021}, we use HEALPix geometry, see https://healpix.jpl.nasa.gov.} and found that $\sim$10 deg$^2$
(HEALPix $N_{side}$=16 and 32, with 3,072 and 12,288 pixels over the full sky) is sufficiently small to capture the
variation of $p(M_r, [Fe/H])$ maps across the sky.
